Rethinking how students learn to design programs

### 2016-04-23 manuscript-body.tex
Finalize Manuscript - RC1

\section{Analyzing only the code Rebecca turned in: What an instructor might see}\label{analyzing-only-the-code-rebecca-turned-in-what-an-instructor-might-see}

@@

Curiously, we have evidence \emph{outside} of Rebecca's code that suggests she knew, and even perhaps had seen, an array-based design solution. In a file called ``notes.txt'' contained in her project directory, we see the following lines:

@@

(notes.txt file, created March 19, 2012)

@@

\section{Analysis Augmenting Code Snapshots with Interview Data}\label{analysis-augmenting-code-snapshots-with-interview-data}

@@

Rebecca's choices become easier to understand when you consider what she said in the interview. When Rebecca initially saw the flights database project, her first reaction was ``this is just a lot like our fantasy football project that we did last year'' (Interview, March 16, 2012).\footnote{Transcript conventions are shown in Appendix 1.} But, the previous fantasy football project was easier because all of the information she needed was in one file. Flights database, by comparison, fractured necessary information across multiple input files.

Since Rebecca said the fantasy football project was like an easier version of flights database, I asked whether she tried making this project more like the easier one. Her response was an emphatic ``Oh yeah! Definitely!'' (Interview, March 16, 2012). She said ``as soon as we got this project I was like, ah! fantasy football! I'm just gonna go and see how much code I can rework from that and like, use /mmhmm/ in this project'' (Interview, March 16, 2012). Specifically, Rebecca went back to her fantasy football code ``and I was looking at how I scanned in the information from the files, cuz, we haven't really done anything like that this semester. Uh, scanning in from files before---'' (Interview, March 16, 2012). Since the fantasy football draft project was the last in which she'd needed to do file-scanning, she ``just, uh, went back to check how I did that /mmhmm/ and then, if I could I copied, but because a lot of the variables were different, uh, like these were more var---less, less variables, and more strings than last year /mmhmm/ uh, I just retyped it out. I just looked at how it was similar'' (Interview, March 16, 2012).

My first opportunity to discuss Rebecca's one-stop flight work was on March 16, 2012, in what would be her third of five interviews that semester. This interview was conducted very early into the time window for the Flight Database project, before Rebecca had done the bulk of her coding. We were discussing her prospective design plans. As Rebecca began explaining how the logic for a one-stop flight search was supposed to work, she described what she saw as one of the central difficulties of the project: the relevant information for answering user queries was spread across multiple files (Interview, March 16, 2012).

@@

Rebecca: Um, like, cuz when we first got this project, uh, I actually was thinking ``oh, well this is just a lot like our fantasy football project we did last year.'' /Huh/ We uh, had to scan in, uh, someone had to enter like ``I wanna pick a quarterback,'' so then you had to scan in and go and look for all the quarterbacks in the file and say ``OK, this is the quarterback'' and everything. But, in that project we only had the one file that had everyone listed: quarterbacks, runningbacks, wide receiver, in one file. And all the information you needed there /Mmhmm/ So, you could just get it all and compare it all at once with one scanf /mmm/ whereas this you have to, take, uh, you scan in the flights, ah, the flights file. So, then you find the flight number. You have to save the ID from that flight number, use that ID to scan into the routes file /mmhmm/ and then save the routes information and then print it out with the flights information. (Interview, March 16, 2012)

@@

Rebecca: That was like, as soon as we got this project I was like, ah! fantasy football! I'm just gonna go and see how much code I can rework from that and like, use /mmhmm/ in this project. And, my whole main file, like all those NULL checks and everything, I mean they're really simple to write, but I just copied `em and put `em there, cuz, we had the same thing. /Mmhmm/ Um, just changed, like, the names of the files.

@@

Rebecca: Because \[reading in from files\] was a big \[Basic Programming\] topic we hadn't talked about it much. /Yup/ So, I just, uh, went back to check how I did that /mmhmm/ and then, if I could I copied, but because a lot of the variables were different, uh, like these were more var---less, less variables, and more strings than last year /mmhmm/ uh, I just retyped it out. I just looked at how it was similar. (Interview, March 16, 2012)

I started doing the scanning, and then I realized like, ``I do not have all the information I need in this one file,'' so that's when I realized I would have to make the dummy variable to compare to the other file, so, which, there was no other file in the, um, fantasy football one. So that's when I realized, I was like OK, so, there's gonna be other stuff that I'm gonna have to do, and /mmhmm/ I mean, I can still kinda compare cuz it's still scanning files /yup/, but the idea of the dummy variable and all that, that would not have come from fantasy football, so /sure/.

@@

Rebecca: So, the way I did it was really long and probably, there was probably like a much easier way, but I just did a giant if---if statements \{swings cursor from line 50 to line 63\} If they wanted to fly on Monday /OK/ I went through and checked to see if the route ID was the same \{wiggles cursor across line 54\}, and if it did, I went through to che---uh, I made a check\_days function \{wiggles cursor across line 60\} uh, I ended up commenting that out cuz I didn't end up /mmhmm/ finishing it. But, uh, my check\_days function worked, it just didn't work completely with the code /OK/ (Interview, April 6, 2012)

@@

Rebecca: And this is why my code, I feel like, is not uh, concise enough, or, I don't really, I forget the word they use, but uh /\{inaudible\}/ it's very long because I couldn't figure out if I should do a while loop or whatever /Uh-huh/ But, so I was just like, I know this way should work if I get everything else right, that uh, just go through, if input's 1, if input's 2 and just do the same thing in each of `em just /Mmmhmm/ check for, ``oh, if days is 2, if days is 1'' instead of, like---Cuz I probably could have done, like, maybe a giant while loop, um, to try and, and if, while, inputs something, uh, then you check to see whatever i is. But, I could, I didn't---couldn't figure out how that would work, so I just did the same thing six times.

@@

Rebecca: Yeah. So that's why it's prob---it's not, uh, the neatest code or whatever, because it's the same thing six times. (Interview, April 6, 2012)

@@

Rebecca: Um, first I'd try and get it to make sure it worked completely /Ahh, OK, yeah/ this way, \{laughs\}, uh, and then, if I had the week after that whatever, I'd probably go through and see if I could figure out a way to make it concise-r because he likes uh, neat, as, like, code that's, uh, easy for the user to see /uh-huh/ I guess. Uh, I forget what, I keep forgetting what the word he used was at the beginning of the year, but uh, just very concise and, uh, this is \[a\] very expanded \{laughs\} way of coding, but, it made sense to me at the time and I was just like ``I just want something that makes sense right now.'' /Right/ So, that I can actually work with and have an idea.

@@

Rebecca: Uh, a little. Like, but, a lot of times in \[Basic Programming\] they didn't really mention too much about being concise. They were just like ``if you can do it, do it'' \{laughs\} /OK/ So I usually stuck to what made sense to me /Right/ uh, to turn the projects in. (Interview, April 6, 2012)

@@

\subsection{Acknowledgments}\label{acknowledgments}

>>>>

\section{Analyzing only the code Rebecca turned in}\label{analyzing-only-the-code-rebecca-turned-in-what-an-instructor-might-see}

@@

Curiously, we have evidence \emph{outside} of Rebecca's code that suggests she knew, and even perhaps had seen, an array-based design solution. In a file called ``notes.txt'' contained in her project directory and created March 19, 2012, we see the following lines:

@@

@@

\section{Augmenting Code Snapshots with Interview Data}\label{analysis-augmenting-code-snapshots-with-interview-data}

@@

Rebecca's choices become easier to understand when we consider what she said in  interviews about the code she wrote. My first opportunity to discuss that code was on March 16, 2012, in what would be her third of five interviews that semester. This interview was conducted very early into the time window for the Flight Database project, before Rebecca had done the bulk of her coding. We were discussing her prospective design plans. As Rebecca began explaining how the logic for a one-stop flight search was supposed to work, she described what she saw as one of the central difficulties of the project: the relevant information for answering user queries was spread across multiple files (Interview, March 16, 2012).

@@

\begin{quote}
  Rebecca: Um, like, cuz when we first got this project, uh, I actually was thinking ``oh, well this is just a lot like our fantasy football project we did last year.'' /Huh/ We uh, had to scan in, uh, someone had to enter like ``I wanna pick a quarterback,'' so then you had to scan in and go and look for all the quarterbacks in the file and say ``OK, this is the quarterback'' and everything. But, in that project we only had the one file that had everyone listed: quarterbacks, runningbacks, wide receiver, in one file. And all the information you needed there /Mmhmm/ So, you could just get it all and compare it all at once with one scanf /mmm/ whereas this you have to, take, uh, you scan in the flights, ah, the flights file. So, then you find the flight number. You have to save the ID from that flight number, use that ID to scan into the routes file /mmhmm/ and then save the routes information and then print it out with the flights information. (Interview, March 16, 2012)
\end{quote}

@@

\begin{quote}
  Rebecca: That was like, as soon as we got this project I was like, ah! fantasy football! I'm just gonna go and see how much code I can rework from that and like, use /mmhmm/ in this project. And, my whole main file, like all those NULL checks and everything, I mean they're really simple to write, but I just copied `em and put `em there, cuz, we had the same thing. /Mmhmm/ Um, just changed, like, the names of the files.
\end{quote}

@@

\begin{quote}
  Rebecca: Because reading in from files was a big Basic Programming topic we hadn't talked about it much. /Yup/ So, I just, uh, went back to check how I did that /mmhmm/ and then, if I could I copied, but because a lot of the variables were different, uh, like these were more var---less, less variables, and more strings than last year /mmhmm/ uh, I just retyped it out. I just looked at how it was similar. (Interview, March 16, 2012)
\end{quote}

@@

\begin{quote}
  Rebecca: So, the way I did it was really long and probably, there was probably like a much easier way, but I just did a giant if---if statements \{swings cursor from line 50 to line 63\} If they wanted to fly on Monday /OK/ I went through and checked to see if the route ID was the same \{wiggles cursor across line 54\}, and if it did, I went through to che---uh, I made a check\_days function \{wiggles cursor across line 60\} uh, I ended up commenting that out cuz I didn't end up /mmhmm/ finishing it. But, uh, my check\_days function worked, it just didn't work completely with the code /OK/ (Interview, April 6, 2012)
\end{quote}

@@

\begin{quote}
  Rebecca: And this is why my code, I feel like, is not uh, concise enough, or, I don't really, I forget the word they use, but uh /\{inaudible\}/ it's very long because I couldn't figure out if I should do a while loop or whatever /Uh-huh/ But, so I was just like, I know this way should work if I get everything else right, that uh, just go through, if input's 1, if input's 2 and just do the same thing in each of `em just /Mmmhmm/ check for, ``oh, if days is 2, if days is 1'' instead of, like---Cuz I probably could have done, like, maybe a giant while loop, um, to try and, and if, while, inputs something, uh, then you check to see whatever i is. But, I could, I didn't---couldn't figure out how that would work, so I just did the same thing six times.

@@

  Rebecca: Yeah. So that's why it's prob---it's not, uh, the neatest code or whatever, because it's the same thing six times. (Interview, April 6, 2012)
\end{quote}

@@

\begin{quote}
  Rebecca: Um, first I'd try and get it to make sure it worked completely /Ahh, OK, yeah/ this way, \{laughs\}, uh, and then, if I had the week after that whatever, I'd probably go through and see if I could figure out a way to make it concise-r because he likes uh, neat, as, like, code that's, uh, easy for the user to see /uh-huh/ I guess. Uh, I forget what, I keep forgetting what the word he used was at the beginning of the year, but uh, just very concise and, uh, this is a very expanded \{laughs\} way of coding, but, it made sense to me at the time and I was just like ``I just want something that makes sense right now.'' /Right/ So, that I can actually work with and have an idea.

@@

  Rebecca: Uh, a little. Like, but, a lot of times in Basic Programming they didn't really mention too much about being concise. They were just like ``if you can do it, do it'' \{laughs\} /OK/ So I usually stuck to what made sense to me /Right/ uh, to turn the projects in. (Interview, April 6, 2012)
\end{quote}

\begin{figure}[htbp]
\centering
\includegraphics{media/CodeTimeline.png}
\caption{We developed CodeTimeline to display snapshots in a timeline. Darkened lines reveals how the seven-fold repeated code began as a very simple switch statement but grew and grew in complexity over time.}
\end{figure}

@@

\section{Conclusion}

Our interview with Rebecca corroborates the snapshot evidence that
Rebecca directly copied code from a prior project. Moreover, Rebecca's
behavior of copying her old fantasy football code suggests several
implications.

First, Rebecca's ``design
inertia'' actually traces as far back as the decisions she made during
the first project of her first semester of programming. That is, by
copying code (and particularly file-scanning logic) from an old project,
she was incorporating core functionality that she designed when she
first learned to program. Admittedly, code reuse isn't in and of itself
a problem in software development. \cite{parsons_cognitive_2004} Parson and Saunders for
example, have written on cognitive heuristics that keep professional
software engineers from reusing code, even when reusing and extending
existing software artifacts is the best course of action on a software
project. So, the concern isn't that Rebecca reused code, but rather the
matter of what cognitive dynamics were at play that directed her choice
to reuse that code. In that regard, the constructs of framing and
transfer may help us better understand Rebecca's activity.

Framing, as it has been applied in contexts such as physics education
\cite{elby_epistemological_2010,hammer_resources_2005,scherr_student_2009} and
mathematics education \cite{vandesande_achieving_2012}, concerns how
participants understand the social and intellectual activities in which
they're engaged. Of particular relevance to studying Rebecca is the
notion of \emph{epistemological framing}, which van de Sande and Greeno
summarize as

\begin{quote}
  participants' understanding of kinds of knowledge that are relevant for
  use in their activity and the kinds of knowledge, understanding, and
  information they need to construct to succeed in their activity (e.g.,
  what kind of information would count as a solution to the problem they
  are working on). \cite{vandesande_achieving_2012}
\end{quote}

Rebecca's decision to copy code from fantasy football implicitly
reflects her orientation toward what kinds of knowledge (the course
topic of scanning information from files) are relevant to solving the
problem. More broadly, note that in Rebecca's interview she explains
that her primary objective is to get a solution that works, which stands
in contrast to having a design priority like having a solution that is
elegant, or one that transparently manages complexity. That commitment
again reflects a manifestation of epistemological framing as ``what
would count as a solution,'' where for Rebecca what counts is a solution
that works.

Additionally, Rebecca saw the flight database project as a new instance
of a prior problem: fantasy football. Consequently, she consciously
adapted past solution patterns, because the flights database project
looked like it contained problems she had already solved in previous
code. Rebecca's deliberate reuse of old code can be readily understood
as an example of what Schwartz, Chase, and Bransford \cite{schwartz_resisting_2012} call
``overzealous transfer'':

\begin{quote}
  Of particular concern are situations where students transfer skills,
  knowledge, and routines that are effective for the task at hand but may
  nevertheless be sub-optimal in the long run because they block
  additional learning. We will call this \emph{overzealous transfer}
  (OZT)---people transfer solutions that appear to be positive because
  they are working well enough, but they are nevertheless negative with
  respect to learning what is new. \cite{schwartz_resisting_2012}
\end{quote}

In short, when students overzealously transfer prior knowledge as
Rebecca did, ``they may believe they are doing the right thing, and
without appropriate feedback they cannot know otherwise'' \cite{schwartz_resisting_2012}

In Rebecca's case, the constructs of framing and overzealous transfer
together let us describe why she would have repurposed a solution that
she felt was adequate, even to the exclusion of the topics being taught
in class and the explicit directions in the project brief. By framing
the flights database problem as a new instance of an old problem,
Rebecca treated it as she did the old problem. But, she arguably
transferred \emph{too} \emph{much} of the old code's structure; so much
that she had to introduce even more complexity into her code just to
make the transferred parts work properly under the new constraints. And,
because her framing of the task seemed to privilege a philosophy of ``if
you can do it, do it'' (Interview, April 6, 2012), her primary goals
were to get her program to work, by whatever means she could understand
and trust.

\section{Acknowledgments}\label{acknowledgments}
